---
title: "What is a Dataset?"
author: "Leighton Pritchard"
date: "2021 Presentation"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: false
    number_sections: true
    css: "../assets/css/style.css"
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("DT")
library("readr")
```

# What is a dataset?

It might be an unhelpful definition, but a dataset is a set of data.

Data comes in many forms, but in science we want our data to be well-organised. To organise our data, we typically collect it into *tables* (in the `R` language, these are referred to as `dataframes`, you may also see them called *matrices*). These are two-dimensional data structures, with *rows* and *columns*, and this allows us to deal with two important and distinct concepts about data: *observations* and *variables* (which may also referred to as *features*, or *measurements*)

## Observations

An *observation* is the act of collecting one or more *measurements* of, or from, an "experimental unit." What constitutes an experimental unit depends on the nature of the experiment. For example, if you measured the [BMI]() of each patient in a GP surgery, the experimental unit would be the patient (and the *measurement* would be their BMI). On the other hand, if you were tracking the concentration of a compound in blood samples taken from a single patient over a period of time, each blood sample would be an experimental unit.

## Variables, Features and Measurements

In an experiment, we measure properties of each "experimental unit." The things that are measured are usually referred to as *variables*, or *features*. In the first example above, the BMI would be termed a *variable*; in the second example, the compound concentration would be called a *variable*.

The terms *variable*, *feature* and *measurement* may often be used interchangeably, but they are sometimes used to imply slightly different things:

- *variable*:
  - a property that varies across the experimental units
- *feature*:
  - (i) a property that varies across the experimental units, but is expected to be constant for each of the experimental units (like eye colour)
  - (ii) a generic term for a *variable*
- *measurement*:
  - (i) a generic term for a *variable*
  - (ii) the actual value that was measured
  
The use of these terms is not always consistent in the literature, and can sometimes be confusing.

## Data Tables

By convention, we represent data in *tables* so that *rows* represent *observations*, and *columns* represent *variables*. The interactive widgets below show some example datasets.

### The `iris` dataset

```{r iris_table, echo=FALSE}
datatable(iris)
```

### The Pima Indians Diabetes Dataset

```{r pima_table, echo=FALSE}
pima = read_csv("../assets/data/pima/pima-indians-diabetes.csv", col_names=c("pregnancies", "conc_Glc", "diastolic", "skin_fold", "insulin", "BMI", "diabetes_pedigree", "age", "class"), col_types="iiiiiddif")
datatable(pima)
```

<div id="questions">
1. How many observations are there in the `pima` dataset?
2. How many variables are there in the `iris` dataset?
</div>
