---
title: "Why Do We Do Statistics?"
author: "Leighton Pritchard"
date: "2021 Presentation"
output: 
  bookdown::html_document2:
    toc: true
    toc_float:
      toc_collapsed: false
    number_sections: true
    css: "../assets/css/rmd_style.css"
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("DT")

# BG color for plots - should match .figure and .caption classes in rmd_style.css
figbg = "whitesmoke"
```

# Introduction

<div id="summary">
- Statistics is a science that aims to solve researchers' problems by analysing data
- This workshop aims to help you critically interpret literature and choose approaches for your own research, but does not focus on theory or mathematical formalism
- The workshop is presented in a flipped style.
  - You should read through the materials, attempt the exercises and optional interactive sections, and bring questions for discussion in the live online session.
- You will need to use `RStudio` Desktop, installed on your own computer, for the optional interactive sections
  - **The live online session is for course material only. Technical questions about running `RStudio` will *not* be addressed at the live online session. Please email the workshop presenters beforehand when you encounter technical issues.**
  - For interactive sections you will be asked to:
    - decide which methods are appropriate for the given problem
    - apply the methods
    - interpret the answers you get
</div>

# Why Do We Do Statistics?

*Statistics* is a science that aims to solve problems mathematically, by representing and analysing *data*. Statistics is applicable to any field that collects data, including medicine, biology, chemistry, engineering, and social sciences. The nature of data can be complex, and so there are always some practical questions that can only be answered by the application of statistical methods to that data.

## What Does Statistics Do?

When researchers carry out experiments, they typically collect *data* for that experiment. This data usually involves at least two *measurements* of each of a number of individuals (or *experimental units*). Sometimes "individuals" means actual individual people, but it might also mean individual medications, groups of people, or something else, depending on context.

Often, one of the measurements is treated as a *response variable* - an *outcome*. For instance, if we were to measure pupil dilation as a result of administering a medication, we could treat the amount of dilation as the response variable. In terms of experimental design, the researcher is attempting to determine how the response variable depends on - or is *explained by* the other measured variables. The response variable is sometimes called the *dependent variable*.

The measured variables that could explain the behaviour of the response variable (i.e. those variables on which the dependent variable might depend) are called *explanatory variables*. For instance, the pupil dilation in our example might depend on: the amount of drug given; the sex of the patient; the age of the patient; and so on. Explanatory variables in an experiment might be measured (i.e. properties of the individual that are observed, like patient sex) or assigned (e.g. giving a drug or a placebo to different groups of individuals).

```{r plant-growth, results="asis", echo=FALSE, fig.align="center"}
datatable(PlantGrowth, width=400,
          caption="Table 1: Data from an experiment assessing the effect of two treatment conditions on yield. `weight` is the response variable, `group` is the (assigned) explanatory variable.")
```

### Understanding and Prediction

Statistics enables two key types of researchers' questions to be answered: *understanding* questions, and *prediction* questions.

- *understanding* questions might take the form of determining whether changing the value of an *explanatory* variable affects the value of the *response* variable. The question might be asking whether there is a change at all, or how much of a change there is.
- *prediction* questions also need to determine a similar relationship between *explanatory* and *response* variables, but they then seek to use those relationships to *predict* responses of individuals on the basis of their measured explanatory variables.

### Models

In all questions, whether for understanding or prediction, we rely on *modelling* the behaviours of the variables, and also their interactions - such as the way that explanatory variables result in the response variable value. Modelling is a broad, complex topic, but for the purpose of this workshop we can consider models to be simplified representations of data, and of the relationships between data. In this workshop, we use two main kinds of model.

Firstly, although we measure data in our experiments, for some statistical approaches we do not use the data directly to calculate output. Instead, we use the data to find *parameters* of a *distribution* (e.g. *mean* and *standard deviation*), such as a Normal or Poisson distribution. This distribution is an idealised *model* of how the data behaves, and it simplifies our approach to use this model instead of the actual data. A key concern in statistics is ensuring that this model is appropriate.

```{r model-normal, echo=FALSE, fig.cap="Two hundred measured data values (histogram) and the corresponding Normal distribution model (curve) of those values. The modelled distribution is entirely described by the parameters: mean=10 (µ) and standard deviation=3 (σ). Note that real data value frequencies often depart from the model; the model simplifies our representation of the 'noisy' data. The dashed line represents the mean value of the model, and the dotted lines represents values at +/- one and two standard deviations from the mean."}
mu = 10
sd = 3

par(bg = figbg)

# "measured" data
data <- rnorm(200, mean=mu, sd=sd)
hist(data, col="cornflowerblue", breaks=40, freq=FALSE,
     main=NULL, xlab="value", ylab="frequency",
     xlim=c(0, 20), ylim=c(0,0.23))

# Curve of distrubution
x = seq(0, 20, length=100)
y = dnorm(x, mean=mu, sd=sd)
lines(x, y)
segments(c(4, 7, 13, 16), c(0, 0, 0, 0), c(4, 7, 13, 16), c(0.2, 0.2, 0.2, 0.2),
         lwd=2, lty=3, col="goldenrod")
segments(10, 0, 10, 0.2, lty=5, lwd=3, col="darkorange1")

text(x=10, y=0.21, expression(mu), col="darkorange4")
text(x=4, y=0.21, expression(mu - 2 * sigma), col="darkorange4")
text(x=7, y=0.21, expression(mu - sigma), col="darkorange4")
text(x=13, y=0.21, expression(mu + sigma), col="darkorange4")
text(x=16, y=0.21, expression(mu + 2* sigma), col="darkorange4")
```

Secondly, the relationship between explanatory and response variables is described by a simplified representation of that relationship: this is also a *model*. For example, if we fit a linear regression between measured values of "plant height" (response variable) and "hours of sunlight received" (explanatory variable), the line of best-fit is a *model* of the relationship between those variables. The inferred values of *intercept* and *gradient* (or *slope*) for this line are *parameters* of that model.

Much of statistical analysis is about estimating the values of parameters for these models, and exploring the implications of the resulting models.

## Why Are We Doing This Workshop

The emphasis of this workshop is to introduce common statistical tools, and help you develop an insight into how they work and when they should, and perhaps also when they ought not to, be applied. This is not a formal mathematical presentation of statistical methods, or statistical *theory*. Rather, this workshop is a primer that describes how and why particular statistical methods are applied to answer research questions, and how to critically interpret their application and interpretation in the literature.

Although we will not dive into theory or mathematics, you should gain an understanding that these methods are thoroughly grounded in mathematics and are not applied in an *ad hoc* way.

## What You Should Expect From This Workshop

The workshop is presented in a *flipped* format. That is, the homework comes *before* the class, and we use an online Zoom session to work though questions and discussions prompted by the material.

This material is broken up into notebooks (like this one) which you should read. Each notebook should work as a standalone piece of material (though they should be read more-or-less in order). You should attempt the exercises in each notebook and - optionally - also attempt the interactive sessions. 

The interactive sessions either fall at the end of a notebook, or are standalone notebooks. Both require that you install [`RStudio` Desktop](https://rstudio.com/products/rstudio/download/) - a free IDE (Integrated Development Environment) on your own computer. The implementation of statistical methods is handled almost exclusively by computers, these days, and `R` is the most common environment for statistical analysis in scientific research. Gaining some familiarity with `R` will help your research, and maybe even your future employability.

<div id="warning">
Please be aware that technical questions about `RStudio` etc. will not be addressed in the live online session. If you encounter technical difficulties you should email one of the workshop presenters when you encounter the problem.
</div>

You will not be expected to write your own *code* in `R`, only to understand the methods, decide which are appropriate to answer the questions you are given, apply the methods, and interpret the answers you get.