---
title: "Where Does the Normal Distribution Come From?"
author: "Leighton Pritchard"
date: "2021 Presentation"
output:
  bookdown::html_document2:
    css: css/rmd_style.css
    theme: lumen
    toc: yes
    toc_float:
      toc_collapsed: no
    number_sections: yes
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)

# BG color for plots - should match .figure and .caption classes in rmd_style.css
figbg = "whitesmoke"
```

<div id="summary">
- Like most statistical distributions, the Normal distribution represents the outcome of a process, equivalent to a real-world activity
- The process used to generate a normal distribution is:
  - Create a large dataset of (continuous) values; this has a (true) mean
  - Repeatedly subsample that set of values, and calculate the mean of each subsample
  - The means of the subsamples are estimates of the *true* mean
- The distribution of subsample means is a Normal distribution
- **The Normal distribution corresponds to repeated estimations of the same (true) continuous variable**
</div>

# Introduction

In most introductory treatments of statistics there is little focus on how statistical distributions 

# Generating a Normal Distribution

## Create a dataset of values

To begin with, we generate a dataset of numbers. These do not themselves have to be normally-distributed. Here, for example, we will generate a number of randomly-chosen values between -10 and +10.

```{r base-samples, echo=FALSE}
inputPanel(
  sliderInput("n_points",
              "Number of datapoints:",
              min = 200,
              max = 1000,
              value = 500),
  
    sliderInput("min_max",
              "Range of datapoints:",
              min = -1000,
              max = 1000,
              value = c(-200, 200)),
  
  actionButton("resample", "Resample")
)
  
dist_data = eventReactive(c(input$resample, input$n_points, input$min_max),
                         data.frame(vals=runif(input$n_points, input$min_max[1], input$min_max[2])),
                         ignoreNULL=FALSE)

renderPlot({
        n_points = input$n_points
        data = dist_data()
        mu = mean(data$vals)
        std = sd(data$vals)
  
        breaks = seq(mu - 3 * std, mu + 3 * std, by=max(1, std/8))      # set up the breakpoints between bars in the histogram
        
        ytop = 1.3 * max(density(data$vals)$y)
        
        p = ggplot(data, aes(x=vals))                                    # set up the ggplot with data
        p = p + geom_histogram(aes(y=..density..),                           # add a historgram layer
                               breaks=breaks,
                               fill="cornflowerblue")
        p = p + annotate("segment",                                          # show the mean as a dashed line
                         x=mu, xend=mu, y=0, yend=0.9 * ytop,
                         colour="darkorange1", size=1, linetype="dashed")
        p = p + annotate("segment",                                          # show standard deviations as dotted lines
                         x=c(mu-2*std, mu-std, mu+std, mu+2*std),
                         xend=c(mu-2*std, mu-std, mu+std, mu+2*std),
                         y=c(0), yend=c(0.9 * ytop),
                         colour="goldenrod", size=1, linetype="dotted")
        p = p + annotate("text",                                             # annotate the lines
                         x=c(mu-2*std, mu-std, mu, mu+std, mu+2*std),
                         y=c(0.95 * ytop),
                         colour="darkorange3",
                         label=c("µ - 2σ", "µ - σ", paste("µ=", format(round(mu, 2), nsmall=2)), "µ + σ", "µ + 2σ"))
        p = p + xlim(-1000, 1000)             # set x-axis limits
        p = p + xlab("measured variable") + ylab("frequency")                # add axis labels
        p = p + theme(plot.background = element_rect(fill = figbg,           # colour background
                                                     color = figbg))
        p    
})

fluidPage(
  mainPanel(
    span("Now we take subsamples from the original dataset")
  )
)

inputPanel(
  sliderInput("n_subsamples",
              "Number of subsamples:",
              min = 3,
              max = 500,
              value = 250),
  
    sliderInput("subsample_size",
              "Datapoints in each subsample:",
              min = 3,
              max = 200,
              value = 25),
  
  actionButton("re_subsample", "Resample")
)

subsamples = eventReactive(c(input$n_subsamples, input$subsample_size, input$re_subsample,
                             input$resample, input$n_points, input$min_max),
                           data.frame(vals=replicate(input$n_subsamples, sample(dist_data()$vals, input$subsample_size))),
                           ignoreNULL=FALSE)

renderPlot({
        means = data.frame(vals=colMeans(subsamples()))
        mu = mean(means$vals)
        std = sd(means$vals)

        breaks = seq(mu - 3 * std, mu + 3 * std, by=max(1, std/8))      # set up the breakpoints between bars in the histogram
        ytop = 1.3 * max(density(means$vals)$y)
        
        p = ggplot(means, aes(x=vals))                                    # set up the ggplot with data
        p = p + geom_histogram(aes(y=..density..),                           # add a historgram layer
                               breaks=breaks,
                               fill="cornflowerblue")
        p = p + stat_function(fun=dnorm, args=list(mean=mu, sd=std),          # add a layer with the Normal curve 
                              geom="line")
        p = p + annotate("segment",                                          # show the mean as a dashed line
                         x=mu, xend=mu, y=0, yend=0.9 * ytop,
                         colour="darkorange1", size=1, linetype="dashed")
        p = p + annotate("segment",                                          # show standard deviations as dotted lines
                         x=c(mu-2*std, mu-std, mu+std, mu+2*std),
                         xend=c(mu-2*std, mu-std, mu+std, mu+2*std),
                         y=c(0), yend=c(0.9 * ytop),
                         colour="goldenrod", size=1, linetype="dotted")
        p = p + annotate("text",                                             # annotate the lines
                         x=c(mu-2*std, mu-std, mu, mu+std, mu+2*std),
                         y=c(0.95 * ytop),
                         colour="darkorange3",
                         label=c("µ - 2σ", "µ - σ", paste("µ=", format(round(mu, 2), nsmall=2)), "µ + σ", "µ + 2σ"))
        p = p + xlim(mu - max(5, 3 * std), mu + max(5, 3 * std))             # set x-axis limits
        p = p + ylim(0, ytop)
        p = p + xlab("measured variable") + ylab("frequency")                # add axis labels
        p = p + theme(plot.background = element_rect(fill = figbg,           # colour background
                                                     color = figbg))
        p    
        
})
```
  
